15yrsChange

### 2024-04-05 doc/15yrsChange.Rmd
update methods pt. 3

In addition to our consideration of the entire forest plot, we focus further on three upland forest portions of the plot (i.e., excluding low-lying areas around streams), with varying levels of exposure to non-indigenous insects and pathogens and white-tailed deer (Fig. 1).  The first portion we define as “Low deer, low canopy vulnerability”: a 4-ha portion of upland forest in the SE quarter of the plot (Fig. **1a**), fenced in 1990 and maintained with only occasional deer presence over the past 24 years. This portion has a low abundance of canopy species vulnerable to non-indigenous pests and pathogens [@holm_interactive_2013]. The second focal area is “high deer, low canopy vulnerability”: 16.3 ha of upland forest area outside the deer exclosure with low abundance of canopy species affected by non-indigenous pests and pathogens (Fig. **1b**). The third portion of the plot is "high deer, high canopy vulnerability": corresponding to 2.2 ha of upland forest outside the deer exclosure with high abundance of canopy species affected by non-indigenous pests and pathogens. We defined upland forest according to topographic wetness index, which was calculated for the forest plot by [@mcgregor_tree_2021a]. We defined vulnerability to non-indigenous insects and pathogens based on the abundance (>20% of tree basal area) of canopy species affected by non-indigenous pests and pathogens at the time of plot establishment in 2008. The following canopy species were identified as vulnerable: ash (*Fraxinus* spp.), elm (*Ulmus* spp.), butternut (*Juglans cinerea*), and American chestnut (*Castanea dentata*) [@anderson-teixeira_longterm_2021].

@@

As part of the ForestGEO network, the plot undergoes a comprehensive woody plant inventory every five years according to the protocol detailed in @condit_tropical_1998. This inventory, hereafter referred to as the census, includes all stems greater than one centimeter in diameter at breast height (1.3 meters; dbh). The census records the dbh, species, living status, and spatial location of each censused stem. Each individual is marked with an identifying number for sequential data collection in subsequent censuses. In multi-stemmed individuals, each additional stem that surpasses the dbh threshold receives a stem number and associated tag. The plot was established in 2008 and recensused three times (2013, 2018, 2023) since, comprising 15 years of forest dynamics data. In the last census, systematic data cleaning was continuously implemented 1) in an ESRI ArcGIS Field Maps application via built-in checks and 2) on a daily basis, via error reports generated through GitHub actions [@kim_implementing_2022]. Each individual’s location was recorded in the Field Maps application using a digitized map of the plot (citation). In addition to the ForestGEO census, a tree health and mortality survey has been conducted annually starting in 2014 on all trees with a dbh ≥ 10 cm in the previous census [@gonzalez-akre_patterns_2016]. Data were collected on the current status, canopy position, percentage of crown living and intact, and visible indicators of tree health, such as physical damage, potential pathogens and insect infestation.

>>>>

In addition to our consideration of the entire forest plot, we focus further on three upland forest portions of the plot (i.e., excluding low-lying areas around streams), with varying levels of exposure to non-indigenous insects and pathogens and white-tailed deer (Fig. 1).  The first portion we define as “Low deer, low canopy vulnerability”: a 4-ha portion of upland forest in the SE quarter of the plot (Fig. **1a**), fenced in 1990 and maintained with only occasional deer presence over the past 24 years. This portion has a low abundance of canopy species vulnerable to non-indigenous pests and pathogens [@holm_interactive_2013]. The second focal area is “high deer, low canopy vulnerability”: 16.3 ha of upland forest area outside the deer exclosure with low abundance of canopy species affected by non-indigenous pests and pathogens (Fig. **1b**). The third portion of the plot is "high deer, high canopy vulnerability": corresponding to 2.2 ha of upland forest outside the deer exclosure with high abundance of canopy species affected by non-indigenous pests and pathogens. We defined upland forest according to topographic wetness index, which was calculated for the forest plot by @mcgregor_tree_2021a. We defined vulnerability to non-indigenous insects and pathogens based on the abundance (>20% of tree basal area) of canopy species affected by non-indigenous pests and pathogens at the time of plot establishment in 2008. The following canopy species were identified as vulnerable: ash (*Fraxinus* spp.), elm (*Ulmus* spp.), butternut (*Juglans cinerea*), and American chestnut (*Castanea dentata*) [@anderson-teixeira_longterm_2021].

@@

As part of the ForestGEO network, the plot undergoes a comprehensive woody plant inventory every five years following the protocol detailed in @condit_tropical_1998. This inventory, hereafter referred to as the census, includes all stems greater than one centimeter in diameter at breast height (1.3 meters; dbh). The census records the dbh, species, living status, and spatial location of each censused stem. Each individual is marked with an identifying number for sequential data collection in subsequent censuses. In multi-stemmed individuals, each additional stem that surpasses the dbh threshold receives a stem number and associated tag. The plot was established in 2008 and recensused three times since (2013, 2018, 2023), providing 15 years of forest dynamics data. In the last census, systematic data cleaning was continuously implemented 1) in an ESRI ArcGIS Field Maps application via built-in checks and 2) on a daily basis, via error reports generated through GitHub actions [@kim_implementing_2022]. Each stem’s location was recorded in the Field Maps application using a digitized map of the plot. In addition to the ForestGEO census, a tree health and mortality survey has been conducted annually starting in 2014 on all trees with a dbh ≥ 10 cm in the previous census [@gonzalez-akre_patterns_2016]. Data were collected on the current status, canopy position, percentage of living and intact crown, and visible indicators of tree health, (*e.g.* physical damage, potential pathogens, insect infestation).
